### 2025-07-29 paper/sections/method.qmd
final pass-through done

The characterization of faithfulness and plausibility in Def. \ref{def-explainability} follows @altmeyer2024faithful, with adapted notation. Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about the input data. Ac tionability constraints in Def. \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we choose to only consider domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We also limit ourselves to classification tasks for reasons discussed in @sec-discussion.

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual. CT adopts gradient-based CE search in training to generate on-the-fly model explanations $\mathbf{x}^\prime$ for the training samples. We use the term *nascent* to denote interim counterfactuals $\mathbf{x}_{t\leq T}^\prime$ that have not yet converged. As we explain below, these nascent counterfactuals can be stored and repurposed as adversarial examples. Conversely, we consider counterfactuals $\mathbf{x}_T^\prime$ as *mature* explanations if they have either exhausted all $T$ iterations or converged by reaching a pre-specified threshold, $\tau$, for the predicted probability of the target class: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.

Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models,
$$
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is any classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and (2) the adversarial loss, $\text{advloss}(.)$, wrt. nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-offs between these components are adjusted through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$. The full training regime is sketched out in Algorithm \ref{alg-experiment}.

@@

To generate samples, @grathwohl2020your use Stochastic Gradient Langevin Dynamics (SGLD) with an uninformative prior for initialization but we depart from their methodology: we propose to leverage counterfactual explainers to generate counterfactuals of observed training samples. Specifically, we have:
$$
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
$$ {#eq-div}
where $\mathcal{E}_\theta(\cdot)$ denotes the energy function defined as $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$, with $y^+$ denoting the index of the randomly drawn target class, $y^+ \sim p(y)$. Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations. Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while increasing the energy of counterfactuals (negative samples) [@du2019implicit]. As the counterfactuals get more plausible (Def. \ref{def-explainability}) during training, these opposing effects gradually balance each other out [@lippe2024uvadlc].

@@

Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@balashankar2023improving;@luu2023counterfactual] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring]. This leaves some flexibility with regards to the choice for the $\text{advloss}(\cdot)$ term in @eq-obj. An intuitive functional form, but likely not the only sensible choice, is inspired by adversarial training:
$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
$$ {#eq-adv}
Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation". Thus, we work with a different distinction between CE and AE than @freiesleben2022intriguing who considers misclassification as the distinguishing feature of adversarial examples. One of the key observations of this work is that we can leverage CEs during training and get AEs essentially for free to reap the aforementioned benefits of adversarial training.

>>>>

The characterization of faithfulness and plausibility in Def. \ref{def-explainability} follows @altmeyer2024faithful, with adapted notation. Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about the input data. Actionability constraints in Def. \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we choose to only consider domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We also limit ourselves to classification tasks for reasons discussed in @sec-discussion.

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual. CT adopts gradient-based CE search in training to generate on-the-fly model explanations $\mathbf{x}^\prime$ for the training samples. We use the term *nascent* to denote interim counterfactuals $\mathbf{x}_{\text{AE}}^\prime$ that have not yet converged. As we explain below, these nascent counterfactuals can be stored and repurposed as adversarial examples. Conversely, we consider counterfactuals $\mathbf{x}_{\text{CE}}^\prime$ as *mature* explanations if they have converged within the $T$ iterations by reaching a pre-specified threshold, $\tau$, for the predicted probability of the target class: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.

Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models,
$$
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_{\text{CE}}^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{\text{AE}}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_{\text{CE}}^\prime,y;\theta)
\end{aligned}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is any classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_{\text{CE}}^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and (2) the adversarial loss, $\text{advloss}(.)$, wrt. nascent counterfactuals $\mathbf{x}_{\text{AE}}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-offs between these components are adjusted through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$. The full training regime is sketched out in Algorithm \ref{alg-experiment}.

@@

To generate samples, @grathwohl2020your use Stochastic Gradient Langevin Dynamics (SGLD) with an uninformative prior for initialization but we depart from their methodology: we propose to leverage counterfactual explainers to generate counterfactuals of observed training samples. Specifically, we have:
$$
\text{div}(\mathbf{x}^+,\mathbf{x}_{\text{CE}}^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_{\text{CE}}^\prime,y)
$$ {#eq-div}
where $\mathcal{E}_\theta(\cdot)$ denotes the energy function defined as $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$, with $y^+$ denoting the index of the randomly drawn target class, $y^+ \sim p(y)$. Conditional on the target class $y^+$, $\mathbf{x}_{\text{CE}}^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations. Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while increasing the energy of counterfactuals (negative samples) [@du2019implicit]. As the counterfactuals get more plausible (Def. \ref{def-explainability}) during training, these opposing effects gradually balance each other out [@lippe2024uvadlc].

@@

Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@balashankar2023improving;@luu2023counterfactual] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring]. This leaves some flexibility with regards to the choice for the $\text{advloss}(\cdot)$ term in @eq-obj. An intuitive functional form, but likely not the only sensible choice, is inspired by adversarial training:
$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{\text{AE}}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
$$ {#eq-adv}
Under this choice, we consider nascent counterfactuals $\mathbf{x}_{\text{AE}}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation". Thus, we work with a different distinction between CE and AE than @freiesleben2022intriguing who considers misclassification as the distinguishing feature of adversarial examples. One of the key observations of this work is that we can leverage CEs during training and get AEs essentially for free to reap the aforementioned benefits of adversarial training.

### 2025-07-29 paper/sections/related.qmd
final pass-through done

CEs have also been used to improve models in the natural language processing domain. For example, @wu2021polyjuice propose *Polyjuice*, a general-purpose CE generator for language models and demonstrate that the augmentation of training data with *Polyjuice* improves robustness in a number of tasks, while @luu2023counterfactual introduce the *Counterfactual Adversarial Training* (CAT) framework that aims to improve generalization and robustness of language models by generating counterfactuals for training samples that are subject to high predictive uncertainty.

>>>>

CEs have also been used to improve models in the natural language processing domain. For example, @wu2021polyjuice propose *Polyjuice*, a general-purpose CE generator for language models, and demonstrate that the augmentation of training data with *Polyjuice* improves robustness in a number of tasks. @luu2023counterfactual introduce the *Counterfactual Adversarial Training* (CAT) framework that aims to improve generalization and robustness of language models by generating counterfactuals for training samples that are subject to high predictive uncertainty.
